\documentclass[11pt]{article}
\input{../preamble.tex}
 
 
\begin{document}
\title{Homework on monoidal categories}
\maketitle

\begin{exercise}
Show that $\fdVec$ and $\Mat$ are equivalent as categories.
\end{exercise}
\begin{proof}[Hint:]
Pick a basis for every finite dimensional vector space.
\end{proof}

\begin{exercise}
Show that every category is equivalent to a skeletal category.
\end{exercise}

\begin{exercise}
Show that $TL(d=1)$ is monoidally equivalent to $\Rep(\bbZ / 2 \bbZ)$.
\end{exercise}

\begin{exercise}
If $\cC$ is a monoidal category, $\cD$ is a category, and we have an equivalence $F:\cC \leftrightarrow \cD:G$, show you can make $D$ into a tensor monoidal in such a way that $F$ and $G$ are a monoidal equivalence.
\end{exercise}
(This is an instance of the idea of `transport of structure`.)

\begin{exercise}
Conclude that every monoidal category is equivalent to a skeletal monoidal category. 
\end{exercise}

\begin{exercise}
Recall that a monoidal functor is a pair: a functor and a natural isomorphism $F(X \otimes Y) \to F(X) \otimes F(Y)$.  Classify all monoidal functors from $\Vec(G)$ to $\Vec(G)$ where the underlying functor is the identity functor. 
\end{exercise}
\begin{proof}[Hint:]
Your answer should involve group cohomology of $G$.
\end{proof}

\begin{exercise}
Find the fusion rules for $TL(q+q^{-1})$ at $q$ a root of unity.

Conclude that the last surviving Jones-Wenzl idempotent is invertible.  This gives a 2-object subcategory where all the objects are invertible --- how does it fit into our classification?
\end{exercise}
\begin{proof}[Hint:]
Suppose we are a root of unity such that $f^{(n+1)}$ is the first negligible Jones-Wenzl idempotent. Prove that $f^{(k)} \otimes f^{(1)} \cong f^{(k-1)} \oplus f^{(k+1)}$ unless $k=0$ or $k=n$. What happens in those cases? From these observations, give a formula for how $f^{(a)} \otimes f^{(b)}$ breaks up as a direct sum of Jones-Wenzl idempotents.
\end{proof}

\begin{exercise}
Show that the golden category, as defined in lecture, is semisimple with two simple objects $1$ and $X$ and in particular that  $X\otimes X \oplus 1 \oplus X$.
\end{exercise}
Recall that the golden category has as objects finite subsets of an interval, and the morphisms are planar trivalent graphs modulo the local relations
\newcommand{\drawI}{ \begin{tikzpicture}[baseline=0cm]
 	\draw (0,.2) .. controls +(30:.3cm) .. (45:.8cm);
 	\draw (0,.2) .. controls +(150:.3cm) .. (135:.8cm);
	\draw (0,.2) -- (0,-.2);
 	\draw (0,-.2) .. controls +(-30:.3cm) .. (-45:.8cm);
 	\draw (0,-.2) .. controls +(-150:.3cm) .. (-135:.8cm);
\end{tikzpicture}
}

\newcommand{\drawH}{ \begin{tikzpicture}[baseline=0cm,rotate=90]
 	\draw (0,.2) .. controls +(30:.3cm) .. (45:.8cm);
 	\draw (0,.2) .. controls +(150:.3cm) .. (135:.8cm);
	\draw (0,.2) -- (0,-.2);
 	\draw (0,-.2) .. controls +(-30:.3cm) .. (-45:.8cm);
 	\draw (0,-.2) .. controls +(-150:.3cm) .. (-135:.8cm);
\end{tikzpicture}}

\newcommand{\cupcap}{\begin{tikzpicture}[baseline=0cm]
	\draw (45:.8cm) .. controls (0,0) .. (135:.8cm);
	\draw (-45:.8cm) .. controls (0,0) .. (-135:.8cm);
\end{tikzpicture}}

\newcommand{\twostrandid}{\begin{tikzpicture}[baseline=0cm,rotate=90]
	\draw (45:.8cm) .. controls (0,0) .. (135:.8cm);
	\draw (-45:.8cm) .. controls (0,0) .. (-135:.8cm);
\end{tikzpicture}}

\newcommand{\unknot}{
\begin{tikzpicture}[baseline=-0.5ex,scale=0.8]
  \draw (0,0) circle (.6cm);
\end{tikzpicture}
}

\newcommand{\loopvertex}{
\begin{tikzpicture}[baseline=-0.5ex,scale=0.8]
  \draw (0,-0.5)--(0,-0.8);
  \draw ([out angle=150]0,-0.5)..(-0.02,0.6)..(0.02,0.6)..([in angle=30]0,-0.5);
\end{tikzpicture}
}

\begin{align*}
\unknot\; &= d \\ 
    \loopvertex\;&=0\\
\drawI \; & = \; \twostrandid \; -  \frac{1}{d}\; \cupcap
\end{align*}

\begin{exercise}
(Checking every detail here is tedious; use your judgement!)

Show that every monoidal category $\cC$ is monoidally equivalent to the strict monoidal category $\mathsf{List} \cC$.

Here $\mathsf{List} \cC$ has as objects words in the objects of $\cC$, and $$\mathsf{List} \cC([x_1, x_2, \ldots x_n] \to [y_1, y_2, \ldots y_m]) = \cC(x_1 \otimes (x_2 \otimes \cdots \otimes (x_n \otimes 1)) \to y_1 \otimes (y_2 \otimes \cdots \otimes (y_m \otimes 1))).$$ Part of the exercise is to define the tensor product of morphisms in $\mathsf{List} \cC$. The functors between $\cC$ and $\mathsf{List} \cC$ should send $x$ to $[x]$ and $[x_1, x_2, \ldots x_n]$ to $x_1 \otimes (x_2 \otimes \cdots \otimes (x_n \otimes 1))$. You'll need to specify what the functors do on morphisms, and make them into monoidal functors by specifying tensorators. Finally you'll need to show that these functors form an equivalence; you can do this directly, or show one of the functors is fully faithful and essentially surjective.
\end{exercise}

\end{document}